---
title: "re_tidy"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r loading in data, include=FALSE}
library(dplyr)
library(lubridate)
library(tidyr)
library(readr)
library(stringr)
library(purrr)

setwd("/Users/michaelweatherford/Library/Mobile Documents/com~apple~CloudDocs/School Work/Grad School/Year 1/Spring Semester/AGDS 1/AGDS")

soilcarbon<-read_csv("./Data/SoilCarbonData_CLEANED.csv")

soilcarbon$log_response <- log(soilcarbon$mean_increased_CO2/soilcarbon$mean_ambient_CO2)

soilcarbon <- soilcarbon %>%
  mutate(Phase = case_when(Time_years < 3 ~ "early",
                           Time_years >=3 & Time_years <=6 ~ "mid",
                           Time_years > 6 ~ "late"))

soilcarbonSummary <- soilcarbon %>%
  group_by(Phase) %>%
  summarise(mean_ambient_CO2=mean(mean_ambient_CO2),
            mean_elevated_CO2=mean(mean_increased_CO2),
            mean_log_response = mean(log_response))
  
```

```{r showing table, echo=TRUE}
knitr::kable(soilcarbonSummary)
```



What are the data that you are looking at?

What do you expect your analysis to show, what is your hypothesis? 

How should soil organic matter content change under elevated CO2?

Interpret your results after aggregating the data: What do your final numbers mean? Do they support your initial hypothesis? Why so, why not?

Tip: Skim the paper and its references to better understand the data and processes you are looking at!``